---
title: "Maps_Ningaloo_Porites_Project"
author: "Lindsey Kraemer"
date: "2024-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load packages #'packageVersion'
```{r, echo=FALSE}
library(tidyverse) #‘2.0.0’
library(sf) #‘1.0.16’
library(RColorBrewer) #‘1.1.3’
library(ggspatial) #‘1.1.9’
library(ggplot2) #‘3.5.1’
library(cowplot) #‘1.1.3’
library(gstat) #‘2.1.2’
library(sp) #‘2.1.4’
library(gridExtra) #‘2.3’
library(scales) #‘1.3.0’

theme_bw()
```

Read Shapefile
```{r}
#(https://www.naturalearthdata.com/downloads) 
#Cite: Made with Natural Earth. Free vector and raster map data @ naturalearthdata.com
#public domain (https://www.naturalearthdata.com/about/terms-of-use/)
world_map <- read_sf("Shapefiles/ne_110m_admin_0_countries/ne_110m_admin_0_countries.shp")

#World Heritage Sites - get Ningaloo Coast Boundaries
#(https://www.vliz.be/en/imis?module=dataset&dasid=8197)
#license: https://creativecommons.org/licenses/by/4.0/
unesco <- read_sf("Shapefiles/WorldMarineHeritageSites_v2/WorldMarineHeritageSites_v2.shp")
```

Extract Australia
```{r}
australia <- world_map[world_map$ADMIN == "Australia", ]
```

Extract UNESCO Ningaloo Coast
```{r}
ningaloo <- unesco[unesco$FULL_NAME == "Ningaloo Coast", ]
```

### Land definition was coarse from naturalearth data and didn't include Muiron Islands, so I created polygons on google earth engine to make sure that the Muiron Islands are included on maps. 

Read in KML file with Muiron Islands coords and polygon of land
```{r}
#path to KML file
#kml_file1 <- "Muirons.kml"

#read the KML file
#muirons <- st_read(kml_file1)

#drop the Z dimension
#muirons_2d <- st_zm(muirons, drop = TRUE, what = "ZM")

#define the output shapefile path
#shapefile_path1 <- "Shapefiles/Muirons.shp"

#write to shapefile
#st_write(muirons_2d, shapefile_path1)

#read in shapefile
muirons <- read_sf("Shapefiles/Muirons.shp")
```

Read in KML file that I made on Google Earth with study site coordinates
```{r}
#kml_file <- "project_sites.kml"

#list all layers in the KML file
#layers_info <- st_layers(kml_file)
#print(layers_info)

#directory to save the shapefiles
#output_dir <- "~/Shapefiles/Sites"

#read and write each layer
#for (layer_name in layers_info$name) {
  #read the layer
#  layer_data <- st_read(kml_file, layer = layer_name)
  
  #convert 3D points to 2D (drop Z dimension)
#  layer_data_2d <- st_zm(layer_data, drop = TRUE, what = "ZM")
  
  #define the output shapefile path
#  shapefile_path <- file.path(output_dir, paste0(layer_name, ".shp"))
  
  #write the layer to a shapefile
#  st_write(layer_data_2d, shapefile_path)
  
  #print a message
#  cat("Converted layer:", layer_name, "to", shapefile_path, "\n")
#}
```

Read Site Shapefiles
```{r}
#define paths to the site shapefiles
shapefiles <- list(
  "Site 1- BUN" = "Shapefiles/Sites/Site 1- Bundegi.shp",
  "Site 2- VLF" = "Shapefiles/Sites/Site 2- VLF Bay.shp",
  "Site 3- OSP" = "Shapefiles/Sites/Site 3- Osprey Bay.shp",
  "Site 4- TAN" = "Shapefiles/Sites/Site 4- Tantabiddi.shp",
  "Site 5- MUI" = "Shapefiles/Sites/Site 5- Muiron Islands.shp",
  "Site 6- COR" = "Shapefiles/Sites/Site 6- Coral Bay.shp",
  "Site 7- LK/TR" = "Shapefiles/Sites/Site 7- Turquoise Bay_Lakeside.shp"
)

#read the shapefiles
site_layers <- lapply(shapefiles, st_read)

#assign names to each layer
names(site_layers) <- names(shapefiles)
```

Focusing in on study area (Ningaloo Coast)
```{r}
#define the bounding box for the Ningaloo area
xmin <- 113.5
xmax <- 114.5
ymin <- -23.2
ymax <- -21.5

```

Make sure shapefiles & layers align geographically
```{r}
target_crs <- st_crs(4326)

#transform CRS for muirons shapefile
if (st_crs(muirons) != target_crs) {
  muirons <- st_transform(muirons, target_crs)
}

#transform CRS for ningaloo shapefile
if (st_crs(ningaloo) != target_crs) {
  ningaloo <- st_transform(ningaloo, target_crs)
}

#transform CRS for each site layer
site_layers <- lapply(site_layers, function(layer) {
  if (st_crs(layer) != target_crs) {
    layer <- st_transform(layer, target_crs)
  }
  return(layer)
})
```

To add in environmental metadata for sites
```{r}
#read the CSV file
gen_site <- read_csv("General_Site.csv") %>%
  mutate(
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude),
    Site = as.factor(Site)  # Ensure Site is a factor
  ) %>%
  mutate(num_Site = paste(site_ID, Site, sep = " - ")) %>%
  select(c(1:11, 15))

#convert to an sf object
gen_site_f <- st_as_sf(gen_site, coords = c("longitude", "latitude"), crs = 4326)
```

## Now to plot data...

Maps of Individual Sites to better view where samples were collected
```{r}
#define color palette
palette <- brewer.pal(n = 7, name = "Set1")
site_colors <- c(
  "Site 1- BUN" = palette[1],
  "Site 2- VLF" = palette[2],
  "Site 3- OSP" = palette[3],
  "Site 4- TAN" = palette[4],
  "Site 5- MUI" = palette[5],
  "Site 6- COR" = palette[6],
  "Site 7- LK/TR" = palette[7]
)

#define bounding boxes for each site
bbox_list <- list(
  "Site 1- BUN" = c(xmin = 114.155, xmax = 114.172, ymin = -21.862, ymax = -21.852),
  "Site 2- VLF" = c(xmin = 114.167, xmax = 114.18, ymin = -21.799, ymax = -21.786),
  "Site 3- OSP" = c(xmin = 113.828, xmax = 113.837, ymin = -22.245, ymax = -22.24),
  "Site 4- TAN" = c(xmin = 113.966, xmax = 113.985, ymin = -21.908, ymax = -21.895),
  "Site 5- MUI" = c(xmin = 114.325, xmax = 114.36, ymin = -21.692, ymax = -21.668),
  "Site 6- COR" = c(xmin = 113.75, xmax = 113.77, ymin = -23.15, ymax = -23.123),
  "Site 7- LK/TR" = c(xmin = 113.882, xmax = 113.913, ymin = -22.105, ymax = -22.035)
)

#function to create a plot for a general site
create_site_plot <- function(site_name, bbox, num_points) {
  ggplot() +
    geom_sf(data = muirons, fill = "gray80", color = "gray50") +
    geom_sf(data = ningaloo, fill = "lightblue", color = "blue", alpha = 0.1) +
    geom_sf(data = site_layers[[site_name]], aes(fill = site_name), shape = 21, size = 2, alpha = 0.9, color = "black", stroke = 0.2) +
    scale_fill_manual(values = site_colors[site_name]) +
    coord_sf(crs = 4326, xlim = c(bbox["xmin"], bbox["xmax"]), ylim = c(bbox["ymin"], bbox["ymax"]), expand = FALSE) +
    theme_minimal() +
    labs(
      title = paste(site_name, "(n =", num_points, ")"),
      x = "Longitude",
      y = "Latitude",
      fill = "Site"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 90, hjust = 1),
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(),  
      axis.ticks = element_line(color = "black"),  
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      panel.background = element_blank(),  
      panel.border = element_rect(color = "black", fill = NA, size = 1) 
    ) +
    annotation_scale(location = "br", width_hint = 0.25, height = unit(0.2, "cm"), text_cex = 0.6) +
    annotation_north_arrow(location = "tl", which_north = "true", 
                           style = north_arrow_fancy_orienteering, height = unit(1, "cm"), width = unit(1, "cm"))
}

#function to create a plot for Site 6 with a different scale bar position
create_site6_plot <- function(bbox, num_points) {
  ggplot() +
    geom_sf(data = muirons, fill = "gray80", color = "gray50") +
    geom_sf(data = ningaloo, fill = "lightblue", color = "blue", alpha = 0.1) +
    geom_sf(data = site_layers[["Site 6- COR"]], aes(fill = "Site 6- COR"), shape = 21, size = 2, color = "black", stroke = 0.2) +
    scale_fill_manual(values = site_colors["Site 6- COR"]) +
    coord_sf(crs = 4326, xlim = c(bbox["xmin"], bbox["xmax"]), ylim = c(bbox["ymin"], bbox["ymax"])) +
    theme_minimal() +
    labs(
      title = paste("Site 6- COR (n =", num_points, ")"),
      x = "Longitude",
      y = "Latitude",
      fill = "Site"
    ) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 90, hjust = 1),
      panel.grid.major = element_blank(),  
      panel.grid.minor = element_blank(),  
      axis.ticks = element_line(color = "black"),  
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      panel.background = element_blank(),  
      panel.border = element_rect(color = "black", fill = NA, size = 1) 
    ) +
    annotation_scale(location = "bl", width_hint = 0.25, height = unit(0.2, "cm"), text_cex = 0.6) +
    annotation_north_arrow(location = "tl", which_north = "true", 
                           style = north_arrow_fancy_orienteering, height = unit(1, "cm"), width = unit(1, "cm"))
}

#create plots for each site and store them in a list
site_plots <- list()
for (site_name in names(bbox_list)) {
  num_points <- nrow(st_drop_geometry(site_layers[[site_name]]))  #count number of points for each site
  if (site_name == "Site 6- COR") {
    plot <- create_site6_plot(bbox_list[[site_name]], num_points)
  } else {
    plot <- create_site_plot(site_name, bbox_list[[site_name]], num_points)
  }
  site_plots[[site_name]] <- plot
}

print(site_plots)


#save each plot as a page in a PDF
pdf("site_plots.pdf", width = 8.5, height = 11)
for (plot in site_plots) {
  print(plot)
}
dev.off()
```

### Supplementary Figure - Combine the supplementary site plots in a grid
```{r}
combined_plot <- plot_grid(plotlist = site_plots, ncol = 2, align = "v")

#display the combined plot
print(combined_plot)

#save the combined plot as an image
ggsave("combined_site_plots.png", combined_plot, width = 12, height = 18, units = "in")
```

# Mapping Environmental Factors

## Climatological Temp Range

```{r}
#load data
clim_data <- read.csv("clim_range_results_bounding_box.csv")

#define a custom gradient for better contrast
gradient_colors <- c("blue", "yellow", "red") 

#create the plot
plot_clim_range <- ggplot() +
    geom_tile(data = clim_data, aes(x = Lon, y = Lat, fill = Clim_Range)) +
    
    #land and base map layers
    geom_sf(data = muirons, fill = "gray80", color = "gray50") +
    geom_sf(data = ningaloo, fill = "lightblue", color = "blue", alpha = 0.1) +
    
    #labels
    geom_text(data = gen_site, aes(x = longitude, y = latitude, label = Site), 
               size = 3, fontface = "bold") +
    
    #enhanced gradient color scale
    scale_fill_gradientn(
        colors = gradient_colors, 
        name = "Climatological Temperature Range (°C)",  
        labels = scales::number_format(accuracy = 0.01)  
    ) +
    
    #coordinate system and map extent
    coord_sf(crs = 4326, xlim = c(113.7, 114.4), ylim = c(-23.15, -21.7)) +
    
    #theme adjustments
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.title = element_text(size = 8),
        axis.text = element_text(size = 8),
        plot.title = element_text(size = 14, face = "bold", hjust = 0),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "bottom",
        legend.box = "vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8)
    ) +
    annotation_scale(location = "br", width_hint = 0.5) +
    annotation_north_arrow(location = "tl", which_north = "true", 
                           style = north_arrow_fancy_orienteering)

print(plot_clim_range)
```

There is missing data or areas where data isn't available due to limitations in observations or model coverage... None of these areas are overlapping with study sites but for the sake of having complete maps, will use Kriging interpolation to help fill these gaps. 

Kriging Interpolation - predicts values for these missing locations by leveraging spatial relationships in the available data, providing a complete surface for visualization or analysis. Unlike many other interpolation techniques, Kriging provides an estimate of uncertainty (variance) for each prediction. 

Kriging Interpolation:

Building a Kriging model with gstat and predicted interpolated values.
```{r}
#convert the data into a spatial object using 'sp'
coordinates(clim_data) <- ~Lon+Lat
proj4string(clim_data) <- CRS("+proj=longlat +datum=WGS84")

#generate the full grid for interpolation using 'epand.grid'
lon_full <- seq(min(clim_data$Lon, na.rm = TRUE), max(clim_data$Lon, na.rm = TRUE), by = 0.05)
lat_full <- seq(min(clim_data$Lat, na.rm = TRUE), max(clim_data$Lat, na.rm = TRUE), by = 0.05)
full_grid <- expand.grid(Lon = lon_full, Lat = lat_full)

#convert the full grid to a SpatialPointsDataFrame
coordinates(full_grid) <- ~ Lon + Lat
proj4string(full_grid) <- CRS("+proj=longlat +datum=WGS84")

#perform Kriging interpolation
kriging_model <- gstat(formula = Clim_Range ~ 1, data = clim_data, nmax = 10)
kriging_result <- predict(kriging_model, newdata = full_grid)

#convert results to a data frame
kriging_filled <- as.data.frame(kriging_result)

#save the interpolated dataset
write.csv(kriging_filled, "kriging_filled_clim_range_results.csv", row.names = FALSE)
```

Call in model
```{r}
kriging_data_c <- read.csv("kriging_filled_clim_range_results.csv")

#check the structure of the data
str(kriging_data_c)
summary(kriging_data_c)

#define custom gradient colors
gradient_colors <- c("blue", "yellow", "red")

#create the plot with the original bounding box
plot_kriging_clim <- ggplot() +
    geom_tile(data = kriging_data_c, aes(x = Lon, y = Lat, fill = var1.pred)) +
    geom_sf(data = muirons, fill = "gray80", color = "gray50") +
    geom_sf(data = ningaloo, fill = "lightblue", color = "blue", alpha = 0.1) +
    geom_text(data = gen_site, aes(x = longitude, y = latitude, label = Site),
              size = 3, fontface = "bold") +
    scale_fill_gradientn(
        colors = gradient_colors, 
        name = "Temperature Range (°C)",  
        labels = scales::number_format(accuracy = 0.01) 
    ) +
    coord_sf(crs = 4326, xlim = c(113.7, 114.4), ylim = c(-23.15, -21.7)) +
    labs(
        title = "(b)",
        x = NULL,
        y = NULL
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_line(color = "black"),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  ) +
  annotation_scale(location = "br", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         style = north_arrow_fancy_orienteering)

print(plot_kriging_clim)
```

## Warm Season Variability

```{r}
#load the warm_season_results_bounding_box.csv data
warm_season_data <- read.csv("warm_season_results_bounding_box.csv")

warm_season_data <- warm_season_data[warm_season_data$Warm_Season_Var > 0, ]


#define a custom gradient for better contrast
gradient_colors <- c("blue", "yellow", "red")  

#create the plot
plot_warm_season <- ggplot() +
    geom_tile(data = warm_season_data, aes(x = Lon, y = Lat, fill = Warm_Season_Var)) +
    geom_sf(data = muirons, fill = "gray80", color = "gray50") + 
    geom_sf(data = ningaloo, fill = "lightblue", color = "blue", alpha = 0.1) +  
    geom_text(data = gen_site, aes(x = longitude, y = latitude, label = Site),
              size = 3, fontface = "bold") +
  
    #enhanced gradient color scale
    scale_fill_gradientn(
        colors = gradient_colors, 
        name = "Warm Season Variability (°C)",
        labels = scales::number_format(accuracy = 0.01) 
    ) +
    
    #coordinate system and map extent
    coord_sf(crs = 4326, xlim = c(113.7, 114.4), ylim = c(-23.15, -21.7)) +
    
    #theme adjustments
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.title = element_text(size = 8),
        axis.text = element_text(size = 8),
        plot.title = element_text(size = 14, face = "bold", hjust = 0),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 1),
        legend.position = "bottom",
        legend.box = "vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8)
    ) +
    annotation_scale(location = "br", width_hint = 0.5) +
    annotation_north_arrow(location = "tl", which_north = "true", 
                           style = north_arrow_fancy_orienteering)


print(plot_warm_season)
```

Now doing a Kriging interpolation.
```{r}
#load the warm_season_results_bounding_box.csv data
warm_season_data <- read.csv("warm_season_results_bounding_box.csv")

#convert the data into a spatial object
coordinates(warm_season_data) <- ~Lon+Lat
proj4string(warm_season_data) <- CRS("+proj=longlat +datum=WGS84")

#generate a regular grid for interpolation
lon_full <- seq(min(warm_season_data$Lon, na.rm = TRUE), max(warm_season_data$Lon, na.rm = TRUE), by = 0.05)
lat_full <- seq(min(warm_season_data$Lat, na.rm = TRUE), max(warm_season_data$Lat, na.rm = TRUE), by = 0.05)
full_grid <- expand.grid(Lon = lon_full, Lat = lat_full)

#convert the full grid to a SpatialPointsDataFrame
coordinates(full_grid) <- ~Lon+Lat
proj4string(full_grid) <- CRS("+proj=longlat +datum=WGS84")

#perform Kriging interpolation
kriging_model <- gstat(formula = Warm_Season_Var ~ 1, data = warm_season_data, nmax = 10)
kriging_result <- predict(kriging_model, newdata = full_grid)

#convert the results to a data frame
kriging_filled <- as.data.frame(kriging_result)

#save the interpolated dataset
write.csv(kriging_filled, "kriging_filled_warm_season_results.csv", row.names = FALSE)

#later on can load the interpolated dataset for visualisation...
kriging_data <- read.csv("kriging_filled_warm_season_results.csv")

#inspect the structure of the data
str(kriging_data)
summary(kriging_data)

#plot the interpolated results
gradient_colors <- c("blue", "yellow", "red") 

plot_kriging_warm <- ggplot() +
  geom_tile(data = kriging_data, aes(x = Lon, y = Lat, fill = var1.pred)) +
  geom_sf(data = muirons, fill = "gray80", color = "gray50") +  
  geom_sf(data = ningaloo, fill = "lightblue", color = "blue", alpha = 0.1) + 
    geom_text(data = gen_site, aes(x = longitude, y = latitude, label = Site),
              size = 3, fontface = "bold") +
  scale_fill_gradientn(
    colors = gradient_colors, 
    name = "Warm Season Variability (°C)", 
    labels = scales::number_format(accuracy = 0.01) 
  ) +
  
  coord_sf(crs = 4326, xlim = c(113.7, 114.4), ylim = c(-23.15, -21.7)) +
  
    labs(
        title = "(c)",
        x = NULL,
        y = NULL
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_line(color = "black"),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  ) +
  annotation_scale(location = "br", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         style = north_arrow_fancy_orienteering)


print(plot_kriging_warm)
```

## Maximum Monthly Mean

```{r}
#load the warmest month SST data
warmest_month_data <- read.csv("warmest_month_sst_bounding_box.csv")

#define gradient colors
gradient_colors <- c("blue", "yellow", "red")

#plot the warmest month SST
plot_warmest_month <- ggplot() +
  geom_tile(data = warmest_month_data, aes(x = Lon, y = Lat, fill = Warmest_Month_SST)) +
  geom_tile(data = warmest_month_data %>% filter(Mask == 0), aes(x = Lon, y = Lat), fill = "gray90") +
  geom_sf(data = muirons, fill = "gray80", color = "gray50") + 
  geom_sf(data = ningaloo, fill = "lightblue", color = "blue", alpha = 0.1) + 
  geom_text(data = gen_site, aes(x = longitude, y = latitude, label = Site), size = 3, fontface = "bold") +
  scale_fill_gradientn(
    colors = gradient_colors,
    name = "Maximum Monthly Mean (°C)",
    labels = scales::number_format(accuracy = 0.01)
  ) +
  
  #coordinate system and bounding box
  coord_sf(crs = 4326, xlim = c(113.7, 114.4), ylim = c(-23.15, -21.7)) +
  
  #theme adjustments
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_line(color = "black"),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  ) +
  annotation_scale(location = "br", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", style = north_arrow_fancy_orienteering)

print(plot_warmest_month)
```

MMM Kriging interpolation
```{r}
#convert the data into a spatial object
coordinates(warmest_month_data) <- ~Lon+Lat
proj4string(warmest_month_data) <- CRS("+proj=longlat +datum=WGS84")

#generate a regular grid for interpolation
lon_full <- seq(min(warmest_month_data$Lon, na.rm = TRUE), max(warmest_month_data$Lon, na.rm = TRUE), by = 0.05)
lat_full <- seq(min(warmest_month_data$Lat, na.rm = TRUE), max(warmest_month_data$Lat, na.rm = TRUE), by = 0.05)
full_grid <- expand.grid(Lon = lon_full, Lat = lat_full)

#convert the full grid to a SpatialPointsDataFrame
coordinates(full_grid) <- ~Lon+Lat
proj4string(full_grid) <- CRS("+proj=longlat +datum=WGS84")

#perform Kriging interpolation
kriging_model <- gstat(formula = Warmest_Month_SST ~ 1, data = warmest_month_data, nmax = 10)
kriging_result <- predict(kriging_model, newdata = full_grid)

#convert the results to a data frame
kriging_filled <- as.data.frame(kriging_result)

#save the interpolated dataset
write.csv(kriging_filled, "kriging_filled_warmest_month_results.csv", row.names = FALSE)

#load the interpolated dataset for visualization
kriging_data <- read.csv("kriging_filled_warmest_month_results.csv")

#inspect the structure of the data
str(kriging_data)
summary(kriging_data)

#plot the interpolated results
gradient_colors <- c("blue", "yellow", "red")  

plot_kriging_warmest <- ggplot() +
  geom_tile(data = kriging_data, aes(x = Lon, y = Lat, fill = var1.pred)) +
  geom_sf(data = muirons, fill = "gray80", color = "gray50") + 
  geom_sf(data = ningaloo, fill = "lightblue", color = "blue", alpha = 0.1) + 
  geom_text(data = gen_site, aes(x = longitude, y = latitude, label = Site),
            size = 3, fontface = "bold") +
  scale_fill_gradientn(
    colors = gradient_colors, 
    name = "Maximum Monthly Mean (°C)",
    labels = scales::number_format(accuracy = 0.01) 
  ) +
  coord_sf(crs = 4326, xlim = c(113.7, 114.4), ylim = c(-23.15, -21.7)) +
    labs(
        title = "(a)",
        x = NULL,
        y = NULL
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_line(color = "black"),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    legend.position = "bottom",
    legend.box = "vertical",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  ) +
  annotation_scale(location = "br", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true", 
                         style = north_arrow_fancy_orienteering)

print(plot_kriging_warmest)
```

### Supplementary Figure- Combining all three plots
```{r}
#arrange the plots side by side
combined_env_plots <- grid.arrange(
  plot_kriging_warmest,
  plot_kriging_clim,
  plot_kriging_warm, 
  ncol = 3
)

#save the combined plot as an image
ggsave("combined_env_plots.png", combined_env_plots, width = 10, height = 6, units = "in")

#inspect
print(combined_env_plots)
```


## Figure 1

Main map showing variables 
```{r}
library(ggnewscale)
library(shadowtext)
library(scico)

dhw_colors <- c(
  "1-3" = "#bdbdbd",
  "3-6" = "#707070",
  "6-9" = "#252525"
)

plot_kriging_clim <- ggplot() +
  geom_tile(
    data = kriging_data_c,
    aes(x = Lon, y = Lat, fill = var1.pred)
  ) +
 scale_fill_gradientn(
    colours = rev(scico::scico(5, palette = "batlow")),
    values = scales::rescale(c(4.58, 4.9, 5.1, 5.3, 5.5, 5.8, 7.05)),
    limits = c(4.58, 7.05),
    breaks = c(4.6, 5.8, 7.0),
    name = "Temperature Range (°C)",
    labels = scales::number_format(accuracy = 0.1)
  ) +
  ggnewscale::new_scale_fill() +
  geom_sf(data = muirons, fill = "#FFF4E1", color = "grey60") +
  geom_sf(data = ningaloo, fill = NA, color = "#2C7FB8", linewidth = 0.6) +
  geom_point(
    data = gen_site,
    aes(x = longitude, y = latitude, fill = factor(DHW_8)),
    shape = 21,
    size = 5,
    color = "black",
    stroke = 0.8
  ) +
 geom_shadowtext(
    data = gen_site,
    aes(x = longitude, y = latitude, label = Site),
    size = 3.5,
    fontface = "bold",
    colour = "black",
    bg.colour = "white",
    bg.r = 0.12,
    nudge_x = -0.07,
    nudge_y = 0.01
  ) +
    scale_fill_manual(
    values = dhw_colors,
    name = "Number of Accumulated Heat Stress Events (DHW ≥ 8°C-weeks)",
    guide = guide_legend(
      override.aes = list(shape = 21, color = "black", size = 4),
      order = 1
    )
  ) +
  coord_sf(crs = 4326, xlim = c(113.6, 114.5), ylim = c(-23.15, -21.6)) +
  labs(
    x = "Longitude",
    y = "Latitude"
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_line(color = "black"),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold", hjust = 0),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    legend.position = "right",
    legend.box = "vertical",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  ) +
  guides(
    fill = guide_legend(override.aes = list(shape = 21, color = "black", size = 4))
  ) +
  annotation_scale(location = "br", width_hint = 0.5) +
  annotation_north_arrow(
    location = "tl",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  )

print(plot_kriging_clim)
```

Create Fig. 1
```{r}
#define bounding box for Australia inset
australia_xmin <- 112
australia_xmax <- 155
australia_ymin <- -44
australia_ymax <- -10


#create the inset plot without lat and lon
inset_plot3 <- ggplot() +
  geom_sf(data = australia, fill = "#FFF4E1", color = "black", size = 0.3) +
  geom_rect(
    aes(xmin = 112.9, xmax = 115, ymin = -23.9, ymax = -21.1),
    color = "#2C7FB8", fill = NA, size = 0.8
  ) +
  coord_sf(
    crs = 4326,
    xlim = c(australia_xmin, australia_xmax),
    ylim = c(australia_ymin, australia_ymax),
    expand = FALSE
  ) +
  scale_x_continuous(
    breaks = seq(115, 150, by = 10),
    labels = function(x) paste0(x, "°E")
  ) +
  scale_y_continuous(
    breaks = seq(-40, -10, by = 10),
    labels = function(y) paste0(abs(y), "°S")
  ) +
  theme(
  axis.text = element_text(size = 6),
  axis.ticks = element_line(size = 0.2),
  axis.title = element_blank(),
  panel.grid = element_blank(),

  panel.background = element_rect(fill = "white", colour = NA),   # panel
  plot.background  = element_rect(fill = NA, colour = NA),   # WHOLE plot
  panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
)

#add the inset plot
final_plot_with_inset <- ggdraw() +
  draw_plot(plot_kriging_clim) +
  draw_plot(inset_plot3, x = 0.35, y = 0.12, width = 0.19, height = 0.19)
  #x=0.55 if the inset plot should be outside of main map

#save
ggsave("final_plot_fig1.png", final_plot_with_inset, width = 9.5, height = 8, units = "in")

print(final_plot_with_inset)
```

Figure out exact colours used in map for each site tile
```{r}
library(ggplot2)
library(dplyr)

#build the plot so ggplot computes the final tile colours
gb <- ggplot_build(plot_kriging_clim)

#first layer is geom_tile() layer
tile_layer <- gb$data[[1]]

#for each site, find the nearest kriging tile and extract its plotted fill colour
site_hues <- gen_site %>%
  rowwise() %>%
  mutate(
    nearest_tile = which.min((tile_layer$x - longitude)^2 + (tile_layer$y - latitude)^2),
    tile_lon = tile_layer$x[nearest_tile],
    tile_lat = tile_layer$y[nearest_tile],
    tile_fill = tile_layer$fill[nearest_tile]
  ) %>%
  ungroup() %>%
  select(Site, longitude, latitude, tile_lon, tile_lat, tile_fill)

print(site_hues)

#to save and use later
gen_site$krig_hex <- site_hues$tile_fill
```

Color Scheme for temp gradient to carry over into phyloseq -> DESeq2 analysis
```{r}
gradient_colors<- c("#204C5F", "#97883D", "#A28B42", 	"#9F8A41", "#818231", "#F5A582", "#A68C44")
```

